Concurrent Programming with CXL

### 2020-06-21 book/paper.tex
Worked on Semaphores chapter

Consider a variant of the Dining Philosophers (which were clearly Dutch---who
else need two forks to eat a meal?) that we call Italian Philosophers (who only
use one fork each to eat spaghetti---they need the other to speak\footnote{The joke
was provided to me by my dear friend and colleague Lorenzo Alvisi.}).
Again, there are five philosophers.
This time, however, there is a glass sitting in the middle of the table with only
three forks in it.  A philosopher needs one of those forks to eat, but clearly,
no more than three philosophers can eat at a time.

\begin{quote}
\begin{itemize}
\item See if you can come up with a deadlock-free solution using locks.  Represent
    each fork as a lock.
\end{itemize}
\end{quote}

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P}
\index{P}
\index{Procure}
operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V}
\index{V}
\index{Vacate}
operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
semaphores.
Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
using a semaphore.

One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
procures the lock, and \texttt{V} vacates the lock.

@@

\section*{Exercices}
\begin{enumerate}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
Note that it uses a shared variable \texttt{a} containing the array to be
sorted and a shared variable \texttt{todo} containing the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check if the array is sorted and a
permutation of the original array.
\end{enumerate}

>>>>

Consider a variant of the Dining Philosophers (which were clearly Dutch---who
else need two forks to eat a meal?) that we call Italian Philosophers (who only
use one fork each to eat spaghetti---they need the other to speak\footnote{The joke
was provided to me by my dear friend and colleague Lorenzo Alvisi.}).
Again, there are five philosophers.
This time, however, there is a glass sitting in the middle of the table with only
three forks in it.  A philosopher needs one of those forks to eat, but clearly,
no more than three philosophers can eat at a time.
See if you can come up with a deadlock-free solution using locks.  Represent
each fork as a lock.  It's not easy.
It is not hard to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.

Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P}
\index{P}
\index{Procure}
operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V}
\index{V}
\index{Vacate}
operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.
One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
procures the lock, and \texttt{V} vacates the lock.

Figure~\ref{fig:semaphore} shows the \texttt{synch} module implementation of
semaphores.
It represents a semaphore as an integer.
Figure~\ref{fig:dinerssema} shows a solution to the Italian Philosphers problem
using a semaphore.

@@

\section*{Exercices}
\begin{enumerate}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in  shared variable \texttt{a}.
Another shared variable, \texttt{todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check that the array is sorted and
that it is a permutation of the original array.
\end{enumerate}
